Staged Abstract Interpreters

### 2019-09-03 paper/10_intro.tex
typos

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes the program by traversing the abstract syntax tree (AST). The traversing incurs
interpretive overhead, such as the pattern matching on the ASTs and recursive
calls on subexpressions. Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can accumulate
significantly if it runs repeatedly (for example, on libraries).  Second, it is
common that abstract interpreters written in pure languages extensively use
functional encodings of side-effects to implement the semantics of abstract interpretation. 
For example, the list monad can be used to model nondeterminism from collecting
interpreters.
%For example, a collecting interpreter that returns all possible runtime values is
%nondeterministic, which can be modeled by a list monad.  
Although such pure approaches have their own merits, compared with imperative,
stateful implementations, they are significantly slower.  In order to
reconcile these approaches, one must use a programming technique which allows
for high-level abstractions during implementation, but which bypasses the
incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed in
1970s, \citet{futamura1971partial} revealed a close
connection between interpreters and compilers through a hierarchy of
specializations, namely, Futamura Projections. The first Futamura
projection shows that specializing an interpreter to the input program
removes interpretation overhead and yields an equivalent executable
that runs faster. The idea of Futamura projection has been applied to 
other similar performance problems and shown success, examples including
query compilers\cite{DBLP:conf/sigmod/TahboubER18}, PyPy \footnote{PyPy:
\url{http://pypy.org/}}, Truffle \cite{Marr:2015:TVP:2814270.2814275} and etc.
The challenge here: can we apply the same idea to a general abstract interpreter
to make it run faster? To achieve this, we need to specialize a nonstandard
semantics that extensively uses nondeterminism, operates on abstract domains,
and computes fixed-point iteratively over a cache. The solution is not simply obvious.

@@

The first proposed approach to realize Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} lets users explicitly program over
a two-level syntax \cite{Nielson:1992:TFL:130665}, by manually
annotating the binding-times of variables in the program. Then, the MSP system
will check whether these annotations are consistent, and specialize the program
using that information. The staging annotations can be either syntactic (e.g.,
quote and quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

Similar to two-level semantics \cite{NIELSON1989117}, we found
monads provide a good abstraction to hide the detail of abstract
interpretation semantics, and allows us to fit code generation into
the semantics. We use \citet{DBLP:journals/pacmpl/DaraisLNH17}'s
abstract definitional interpreter that uses monad transformers as the
unstaged baseline. However, the monads do not have a clear stage distinction.
We further introduce binding-time and make the monads binding-time polymorphic.
Our proposed framework adopts the type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the Futamura projection
of a big-step abstract interpreter for a small, stateful, higher-order
language. By deriving staged monads that can be used to do code generation, we
obtain a staged abstract interpreter that can be used for specialization. After
staging, the generated code is specialized to the input program, and all
monadic operations are inlined and compiled down to low-level Scala code.

@@

Although the four artifacts may look like dissimilar at first glance, they in
fact are all firmly rooted in the concrete semantics of the language.  This
observation provides a way to abstract over the interpreter and achieve the
flexibility of reinterpreting the shared interpreter.  The generic interpreter
returns a value of monadic type, which can be varied by different semantics.
The domain of the interpreter and the effects such as read, state, and
nondeterminism can all be wrapped into this monadic type.  The binding-time
abstraction is represented by a higher-kinded type, and controls whether the
interpreter produces values directly or generates code. It is worth mentioning
that the binding-time type is also injected into the monadic type, so that we
will distinguish normal monads and staged monads.

\paragraph{Applications and Evaluations}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, an implementation technique for
control-flow analyses, and show that by utilizing type-based stage
annotations we can achieve the same optimizations. Meanwhile,
the analyzer does not need to change, thereby requiring significantly less
engineering effort.
2) We extend the basic staged abstract interpreter to different flow
analyses, including a store-widened analysis, a context-sensitive
analysis, and abstract garbage collection.
3) We show that staging an abstract interpreter enables modular
compilation of an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore, the approach provides a modular way to create optimized
analysis code by mechanized reuse of a whole-program analyzer.
4) We also empirically evaluate the performance improvements gained by staging,
showing an order of magnitude speedup on flow-analysis.

\paragraph{Contributions} Briefly, the contributions of the paper are as follows:
\begin{itemize}[leftmargin=2em]
  \item To our best knowledge, we present the first
    multi-stage programming system to specialize a general abstract interpreter,
    without our approach affecting the soundness of analyses.
  \item Intellectually, our framework naturally extends the first
    Futamura projection to abstract interpreters, showing a
    well-grounded approach to optimizing static analyses via
    meta-programming.
  \item Practically, we show that staging an abstract interpreter is
    useful to improve performance and scalability of analyses by case
    studies and an empirical evaluation.
\end{itemize}

>>>>

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes the program by traversing the abstract syntax tree (AST). The traversing incurs
interpretive overhead, such as the pattern matching on the ASTs and recursive
calls on subexpressions. Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can accumulate
significantly if it runs repeatedly (for example, on libraries).  Second, it is
common that abstract interpreters written in pure languages extensively use
functional encodings of side-effects to implement the semantics of abstract interpretation. 
For example, the list monad can be used to model nondeterminism from collecting
interpreters.
%For example, a collecting interpreter that returns all possible runtime values is
%nondeterministic, which can be modeled by a list monad.  
Although such pure approaches have their own merits, compared with imperative,
stateful implementations, they are significantly slower.  In order to
reconcile these approaches, one must use a programming technique which allows
for high-level abstractions during implementation, but bypasses the
incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed in
1970s, \citet{futamura1971partial} revealed a close
connection between interpreters and compilers through a hierarchy of
specializations, namely, Futamura Projections. The first Futamura
projection shows that specializing an interpreter to the input program
removes interpretation overhead and yields an equivalent executable
that runs faster. The Futamura projections has been successfully applied to 
other similar performance problems, examples including
query compilers\cite{DBLP:conf/sigmod/TahboubER18}, PyPy \footnote{PyPy:
\url{http://pypy.org/}}, Truffle \cite{Marr:2015:TVP:2814270.2814275} and etc.
The challenge here: can we apply the same idea to a general abstract interpreter
to make it run faster? To achieve this, we need to specialize a nonstandard
semantics that extensively uses nondeterminism, operates on abstract domains,
and computes fixed-point iteratively over a cache. The solution is not simply obvious.

@@

The first proposed approach to realize Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} allows users explicitly program over
a two-level syntax \cite{Nielson:1992:TFL:130665}, by manually
annotating the binding-times of variables in the program. Then, the MSP system
will check whether these annotations are consistent, and specialize the program
using that information. The staging annotations can be either syntactic (e.g.,
quote and quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

To apply Futamura projections on abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract
interpreter as the unstaged baseline.  Similar to the two-level semantics
\cite{NIELSON1989117}, monads provide a good abstraction to hide the detail of
abstract interpretation semantics. However, the monads do not have a clear
stage distinction.
We further introduce binding-times and make the monads binding-time polymorphic.
Our proposed framework adopts the type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the Futamura projection
of a big-step abstract interpreter for a small, stateful, higher-order
language. By deriving staged monads that can be used to do code generation, we
obtain a staged abstract interpreter that can be used for specialization. After
staging, the generated code is specialized to the input program, and all the
monadic operations are inlined and compiled down to low-level Scala code.

@@

Although the four artifacts may look dissimilar at first glance, they in
fact are all firmly rooted in the concrete semantics of the language.  This
observation provides a way to abstract over the interpreter and achieve the
flexibility of reinterpreting the shared interpreter.  The generic interpreter
returns a value of monadic type, which can be varied by different semantics.
The domain of the interpreter and the effects such as read, state, and
nondeterminism can all be wrapped into this monadic type.  The binding-time
abstraction is represented by a higher-kinded type, and controls whether the
interpreter produces values directly or generates code. It is worth mentioning
that the binding-time type is also injected into the monadic type, so that we
will distinguish normal monads and staged monads.

\paragraph{Applications and Evaluation}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, an implementation technique for
control-flow analyses. We show that by utilizing type-based stage
annotations, we can achieve the same optimizations. Meanwhile,
the analyzer does not need to change, except the stage annotations, 
thereby requiring significantly less engineering effort.
2) We extend the basic staged abstract interpreter to different flow
analyses, including a store-widened analysis, a context-sensitive
analysis, and abstract garbage collection.
3) We show that staging an abstract interpreter enables modular
compilation of an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore, the approach provides a modular way to create optimized
analysis code by mechanized reuse of a whole-program analyzer.
4) We empirically evaluate the performance improvements gained by staging,
showing an order of magnitude speedup on flow analyses.

\paragraph{Contributions} Briefly, the contributions of the paper are as follows:
\begin{itemize}[leftmargin=2em]
  \item Intellectually, we present a framework that naturally extends the first
    Futamura projection to definitional abstract interpreters, showing a
    well-grounded approach to optimizing static analyses via meta-programming.
  \item Practically, we show that staging an abstract interpreter is useful to
    improve performance and scalability of analyses by case studies and an
    empirical evaluation.
\end{itemize}

### 2019-09-03 paper/60_cases.tex
typos

We borrow the notation traditional modular analysis
\cite{DBLP:conf/cc/CousotC02} and realize that such generated function is a
reusable artifact which can be used modularly, i.e., independent to its context. 
The generated functions already eliminate the interpretation overhead and
monadic overhead, therefore can be consider as partial summaries. Although, we
still need to run the generated function at next stage to obtain the analysis
result.  At next stage, we can provide different abstract arguments, store and
caches, depending on the context where it is called.

>>>>

We borrow the notation from traditional modular analysis
\cite{DBLP:conf/cc/CousotC02}: such generated function is a
reusable artifact which can be used modularly, i.e., independent to its context. 
The generated functions already eliminate the interpretation overhead and
monadic overhead, therefore can be consider as partial summaries. Although, we
still need to run the generated function at next stage to obtain the analysis
result.  At next stage, we can provide different abstract arguments, store and
caches, depending on the context where it is called.
